---
title: "Insurance Forecast for a Medical Cost Personal Datasets"
date: "2025-04-07"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```




>>Source:  https://www.kaggle.com/datasets/mirichoi0218/insurance



```
install.packages("tidyverse")
```

## Including Plots

You can also embed plots, for example:

```{r pressure, echo=FALSE}
library(tidyverse)
library(MASS)
```

Note that the `echo = FALSE` parameter was added to the code chunk to prevent printing of the R code that generated the plot.
```{r}
insurance = read.csv(url("https://raw.githubusercontent.com/stedy/Machine-Learning-with-R-datasets/refs/heads/master/insurance.csv"))

```


```{r}

# Data Preprocessing

# Check structure and summary
str(insurance)
summary(insurance)

# Check for missing values
cat("Missing Values in Each Column:\n")
print(colSums(is.na(insurance)))

#categorical variables are factors
insurance$sex <- as.factor(insurance$sex)
insurance$smoker <- as.factor(insurance$smoker)
insurance$region <- as.factor(insurance$region)

```


```{r}
bmi_category <- cut(insurance$bmi,
                             breaks = c(0, 18.5, 25, 30, Inf),
                             labels = c("Underweight", "Normal", "Overweight", "Obese"))
age_group <- cut(insurance$age,
                           breaks = c(0, 20, 30, 40, 50, 60, Inf),
                           labels = c("0–20", "21–30", "31–40", "41–50", "51–60", "61+"))

```


```{r}
library(dplyr) 
glimpse(insurance)
```


>> Histogram of Smoker and non-smoker

```{r}
table(insurance$smoker)
smoker_age <- insurance$age[insurance$smoker == "yes"]
nonsmoker_age <- insurance$age[insurance$smoker == "no"]


par(mfrow = c(1, 2))

hist(nonsmoker_age, 
     main = "Non-Smokers by Age", 
     xlab = "Age", 
     col = "green", 
     border = "white",
     breaks = 5)

hist(smoker_age, 
     main = "Smokers by Age", 
     xlab = "Age", 
     col = "red", 
     border = "white",
     breaks = 5)

```
```{r}
barplot(table(insurance$sex),
        main = "Count by Sex",
        xlab = "Sex",
        ylab = "Frequency",
        col = c("blue", "red"),
        ylim = c(0,700))
```
>>Overlay Histogram

```{r}
ggplot(data = insurance, aes(x= charges))+
  geom_histogram(aes(fill = sex))
```

>>Distribution of Insurance Charges

```{r}
ggplot(insurance, aes(x = charges)) +
  geom_histogram(binwidth = 2000, fill = "blue", color = "black") +
  labs(title = "Distribution of Insurance Charges",
       x = "Charges", y = "Count") +
  scale_x_continuous(limits = c(0, 60000), breaks = seq(5000, 60000, 5000)) +
  scale_y_continuous(limits = c(0, 150), breaks = seq(0, 200, 25))
```



```{r}
ggplot(insurance, aes(x = smoker, y = charges, fill = smoker)) +
  geom_boxplot() +
  theme_minimal() +
  ggtitle("Insurance Charges by Smoking Status")
```

```{r}
ggplot(insurance, aes(x = sex, y = charges, fill = sex)) +
  geom_boxplot() +
  theme_minimal() +
  ggtitle("Insurance Charges by Male/Female")
```



>>Boxplot of different regions

```{r}
ggplot(data = insurance, mapping = aes(x = region, y = charges)) +
  geom_boxplot()+ ggtitle("Boxplot of Different regions")
```
```{r}
ggplot(insurance, aes(x = charges, fill = smoker)) +
  geom_histogram(alpha = 0.5, position = "identity", binwidth = 2000) +
  labs(title = "Overlayed Histogram of Charges by Smoking Status",
       x = "Charges", y = "Count") +
  scale_fill_manual(values = c("green", "red")) +
  theme_minimal()
```
 >>Charges vs BMI (colored by smoker)

```{r}
ggplot(insurance, aes(x = bmi, y = charges, color = smoker)) +
  geom_point(alpha = 0.6) +
  geom_smooth(method = "lm", se = FALSE) +
  theme_minimal() +
  ggtitle("Charges vs BMI colored by Smoking Status")

```


```{r}
ggplot(insurance, aes(x = bmi_category, y = charges)) +
  geom_boxplot(fill = "lightgreen") +
  labs(title = "Insurance Charges by BMI Category",
       x = "BMI Category",
       y = "Charges") +
  theme_minimal()

ggplot(insurance, aes(x = "", y = bmi)) +
  geom_boxplot(fill = "skyblue") +
  labs(title = "Boxplot of BMI", y = "BMI") +
  theme_minimal()
```



```{r}
ggplot(insurance, aes(x = region, fill = smoker)) +
  geom_bar(position = "dodge") +
  labs(title = "Number of Smokers and Non-Smokers by Region",
       x = "Region",
       y = "Count",
       fill = "Smoker Status") +
  theme_minimal() +
  scale_fill_manual(values = c("no" = "#00BFC4", "yes" = "#F8766D"))
  

```

```{r}
ggplot(insurance, aes(x = bmi_category, y = children, fill = smoker)) +
  geom_boxplot() +
  labs(title = "Children Count by BMI Group and Smoking Status",
       x = "BMI Group", y = "Number of Children") +
  theme_minimal()
```



>> Null hypothesis (H₀): Variance of smoker = variance of non smoker

>>Alternate Hypothesos H1 : Variance of smoker > variance of non smoker

```{r}
smoker_charges <- insurance$charges[insurance$smoker == "yes"]
nonsmoker_charges <- insurance$charges[insurance$smoker == "no"]
var.test(smoker_charges,nonsmoker_charges,alternative = "greater",conf.level = 0.95)
```

>>Since p-value is less than 0.05 , I can conclude that the variances are different and we should do a t-test with different variances.


>>Null hypothesis (H₀): μ1 = μ2 (no difference in average charges)

>>Alternative hypothesis (H₁): μ1 > μ2 (smokers have higher average charges)

```{r}


#one-sided t-test
t.test(smoker_charges, nonsmoker_charges, alternative = "greater",conf.level = .95,var.equal = FALSE)
```
>The p-value is less than 0.05 and from this I have rejected the null hypothesis(there is no difference in the average charges). 


```{r}
library(ggplot2)
ggplot(insurance, aes(x = smoker, y = charges, fill = smoker)) +
  geom_boxplot(alpha = 0.6, outlier.color = "red") +
  stat_summary(fun = mean, geom = "point", shape = 20, size = 4, color = "black", fill = "black") +
  labs(
    title = "Comparison of Insurance Charges by Smoking Status",
    x = "Smoker",
    y = "Charges ($)",
    fill = "Smoker Status"
  ) +
  scale_fill_manual(values = c("no" = "skyblue", "yes" = "tomato")) +
  theme_minimal()
```
>From the ggplot above, it clearly shows that the people who smoke have a higher median insurane charges than people who dont smoke.

```{r}
ggplot(insurance, aes(x = age_group, y = charges)) +
  geom_boxplot(fill = "green", color = "black") +
  labs(title = "Boxplot of Charges by Age Group",
       x = "Age Group",
       y = "Charges") +
  theme_minimal()
```


```{r}

alpha = 0.05
insurance_alpha = qt(1-alpha/2,1337)
insurance_alpha

```




>Anova Testing

>NUll Hypothesis H0 = Average insurance cost are equal across the BMI categories

>Alternate hypothessis H1 -> Atleast one BMI category has a significant different insurance charge.

```{r}


bmi_anova <- aov(charges ~ bmi_category, data=insurance)


summary(bmi_anova)

table(insurance$bmi_category)

```
>>Since the p-value is less than the sigificance level of 0.05 , we reject the null hypothesis.


```{r}
library(ggplot2)
ggplot(insurance, aes(x=bmi_category, y=charges, fill=bmi_category)) +
  geom_boxplot(alpha=0.7) +
  stat_summary(fun=mean, geom="point", shape=23, size=3, color="black") +
  labs(title="Medical Charges by BMI Category",
       x="BMI Category",
       y="Charges ($)") +
  scale_fill_brewer(palette="Set2") +
  theme_minimal()
```



>>Linear Regression

>> We have compared linear regression by comapring with smoker , bmi and smoker plus the bmi category seperately.


```{r}
insurancelmsmoking = lm(charges~smoker,data = insurance)
summary(insurancelmsmoking)


insurancelmbmi = lm(charges~bmi , data = insurance)
summary(insurancelmbmi)


#plot(insurance$bmi,insurance$charges)
#abline(insurancelmbmi,col="red")

lm1 = lm(charges~smoker+bmi_category , data = insurance)
summary(lm1)

lm2 = lm(charges~age_group,data = insurance)
summary(lm2)


```




>>Multiple Linear Regression

```{r}

# Splitting the data into 80% training and 20% testing

i = sample(2,nrow(insurance), replace = TRUE,prob = c(0.8,0.2) )
insuranceTraining = insurance[i==1,]
insuranceTest = insurance[i==2,]


#pairs(insuranceTraining[,1:7],lower.panel = NULL)
```

>>Full Model

```{r}
fullModel = lm(charges~., data = insuranceTraining[,1:7])
summary(fullModel)
```




```{r}

#Predicting the fullModel with insuranceTest


library(MLmetrics)
ypred <- predict(fullModel,insuranceTest)
summary(ypred)
# Evaluate the model
fullModel_MAE = MAE(ypred, insuranceTest$charges)
fullModel_MSE = MSE(ypred, insuranceTest$charges)
  
#mean((ypred - insuranceTest$charges)^2)
#rmse <- sqrt(mse)
#cat("\nRoot Mean Squared Error (RMSE):", rmse, "\n")
cat("\n MAE is : ",fullModel_MAE)
cat("\n MSE is : ",fullModel_MSE)
#rmse = sqrt(fullModel_MSE)



#rmse

  
```


```{r}
#par(mfrow = c(2,2))
#plot(fitlm)
```


>>Forward Stepwise Model


```{r}


intercept_only = lm(charges~1, data = insuranceTraining)

all = lm(charges~. , data = insuranceTraining[,1:7])

forward = stepAIC(intercept_only,direction = 'forward',scope = formula(all))

```
```{r}
# view results of forward stepwise regression
forward$anova
```
```{r}
# view final model
summary(forward)
```




```{r}
#Prediction using the forward stepwise model

library(MLmetrics)
ypred_forward <-predict(object=forward,newdata = insuranceTest[,1:7])
summary(ypred_forward)

forward_MAE=MAE(ypred_forward,insuranceTest$charges)
forward_MSE=MSE(ypred_forward,insuranceTest$charges)

cat("\n MAE is : ",forward_MAE)
cat("\n MSE is : ",forward_MSE)
```

>> Backward Stepwise 

```{r}
backward = stepAIC(all,direction = 'backward')

```
```{r}
# view results of backward stepwise regression
backward$anova
```
```{r}
summary(backward)
```



#Prediction using backward

```{r}
ypred_backward <-predict(object = backward, newdata = insuranceTest[])
summary(ypred_backward)
MAE(ypred_backward,insuranceTest$charges)
MSE(ypred_backward,insuranceTest$charges)
```
```{r}
summary(fullModel)
summary(forward)
summary(backward)
```



>>Random Forest


```{r}

library(randomForest)
rf_model <- randomForest(charges ~ ., data = insuranceTraining)
predictions <- predict(rf_model, newdata = insuranceTest)
sqrt(mean((predictions - insuranceTest$charges)^2))

```

```{r}
plot(insuranceTest$charges, predictions,
     xlab = "Actual Charges",
     ylab = "Predicted Charges",
     main = "Actual vs Predicted (Random Forest)",
     col = "blue", pch = 16)
abline(0, 1, col = "red")
```


>>Xg Boost

```{r}
# Install and load necessary package

library(xgboost)

# Prepare the data for XGBoost
train_matrix <- data.matrix(insuranceTraining[, -which(names(insuranceTraining) == "charges")])  # Exclude target variable
test_matrix <- data.matrix(insuranceTest[, -which(names(insuranceTest) == "charges")])

train_labels <- insuranceTraining$charges
test_labels <- insuranceTest$charges

# Convert data into DMatrix format
dtrain <- xgb.DMatrix(data = train_matrix, label = train_labels)
dtest <- xgb.DMatrix(data = test_matrix, label = test_labels)

# Set parameters for the XGBoost model
params <- list(
  objective = "reg:squarederror", # Regression task
  booster = "gbtree", # Tree booster
  eta = 0.1, # Learning rate
  max_depth = 6, # Maximum depth of trees
  colsample_bytree = 0.8, # Fraction of features to consider per tree
  subsample = 0.8 # Fraction of samples to use per tree
)

# Train the XGBoost model
xgb_model <- xgb.train(params = params, data = dtrain, nrounds = 100)

# Make predictions on the test set
predictions_xgb <- predict(xgb_model, newdata = dtest)

# Evaluate the model with RMSE
mse_xgb <- mean((predictions_xgb - test_labels)^2)
rmse_xgb <- sqrt(mse_xgb)
cat("\nRoot Mean Squared Error (RMSE) for XGBoost:", rmse_xgb, "\n")

# Plot Actual vs Predicted values
library(ggplot2)
results <- data.frame(Actual = test_labels, Predicted = predictions_xgb)
ggplot(results, aes(x = Actual, y = Predicted)) +
  geom_point(color = 'blue') +
  geom_abline(intercept = 0, slope = 1, color = 'red', linetype = "dashed") +
  labs(title = "Actual vs Predicted (XGBoost)", x = "Actual Values", y = "Predicted Values") +
  theme_minimal()

```

